Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-12-04 workflow/workflow.Rmd
comments

EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'

@@

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
core <- dataObj[names(vars)[1:1000],]
```

@@

EP: Why do we create 'core' rather than just replace 'dataObj'

@@

```{r}
core
```

@@

```{r batch}
batch <- colData(core)$Batch
col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
              brewer.pal(8, "Accent")[1])
names(col_batch) = unique(batch)
table(batch)
```

@@

```{r original}
publishedClusters <- colData(core)[, "publishedClusters"]
col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
names(col_clus) <- sort(unique(publishedClusters))
table(publishedClusters)
```

@@

```{r zinb}
print(system.time(dataObj <- zinbwave(core, K = 50, X = "~ Batch",  
                                 residuals = TRUE,
                                 normalizedValues = TRUE)))
```

@@

EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization. The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought. Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.

@@

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- colData(se)[, grepl("^W", colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = "Sample")
```

@@

EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.

@@

EP: Currently isCount=FALSE. Does that make sense for the normalized values, or should we set it to TRUE (it shouldn't affect the clustering, since the clustering is only on the reducedDim slot now, but could affect downstream plots, etc.).

EP: We could add back in the mergeClusters step now that we don't have the W in the data slot. I haven't done that yet because we want to just recreate the previous, but it would be reasonable now. It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously. But then again, I don't know the workflow for the DE part, so maybe we want to leave it out.  But it would be nice to show that you can do that for this setting. Once this workflow is running I can come back to it.

>>>>

EP: changing 'se' to 'dataObj' because it changes classes. Also can search and replace 'dataObj' and not 'se'.  
FP: good idea.

@@

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
dataObj <- dataObj[names(vars)[1:1000],]
```

@@

EP: Why do we create 'core' rather than just replace 'dataObj'. I agree, I think that it is unnecessary. I changed 'core' to dataObj.

@@

```{r}
dataObj
```

@@

```{r batch}
batch <- colData(dataObj)$Batch
col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
              brewer.pal(8, "Accent")[1])
names(col_batch) = unique(batch)
table(batch)
```

@@

```{r original}
publishedClusters <- colData(dataObj)[, "publishedClusters"]
col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
names(col_clus) <- sort(unique(publishedClusters))
table(publishedClusters)
```

@@

```{r zinb}
print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch", residuals = TRUE, normalizedValues = TRUE)))
```

@@

EP: Should probably address these normalized values, especially to more strongly make clear that shouldn't use these for clustering, just for visualization. The way its stated it sounds like dimensionality reduction is an "also", when in fact it is the main point and the normalized values are the afterthought. Might consider reordering the sections to more clearly demonstrate this and put dimensionality reduction first.  
FP: Agree, let's keep it in mind for once the workflow runs smoothly.

@@

FP: W can now be accessed using reducedDim function. Two functions called reducedDim are loaded in the workflow: one from slingshot and the other one from singleCellExperiment.

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- SingleCellExperiment::reducedDim(dataObj)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, fill = unique(col_clus), title = "Sample")
```

@@

EP: If zinbwave already has been updated, this above code is probably not needed since its aready there.  
FP: As the name/type "zinbwave50" is then needed in the RSEC function, we might need to keep this line.

@@

EP: Currently isCount=FALSE. Does that make sense for the normalized values, or should we set it to TRUE (it shouldn't affect the clustering, since the clustering is only on the reducedDim slot now, but could affect downstream plots, etc.).  
FP: yes, I think isCount=FALSE makes sense for the normalized values.

EP: We could add back in the mergeClusters step now that we don't have the W in the data slot. I haven't done that yet because we want to just recreate the previous, but it would be reasonable now. It would be based on doing DE of the normalized values, however, because I put them into the main assay slot, so we might need to think about this multiple assay question more seriously. But then again, I don't know the workflow for the DE part, so maybe we want to leave it out.  But it would be nice to show that you can do that for this setting. Once this workflow is running I can come back to it.  
FP: I am not sure it makes sense to perform the DE on the normalized values. Intuitively, I would say no, but not sure. We can still show that it is a possibility and write in the text that in this case, it might not make sense. We are working on a DE method for scRNA-seq where we use observational weights computed from the posterior probabilities of the ZINB-WaVE model in bulk RNA tools like edgeR and DESeq2. I think, eventually, it would be nice to be able to use that. Maybe, not for the next next version of the workflow.
